---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Will Thompson"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)

F_0  <- 73.00
r    <- 0.04
mult <- 1000
  
port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

+ Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
+ You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
+ Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
+ No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1. Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range ($50–$100/bbl).
2. Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3. Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4. Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

+ **Delta** (∂V/∂F)
+ **Gamma** (∂²V/∂F²)
+ **Vega** (∂V/∂σ)
+ **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in $/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = $1,500,000 per $1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ [50, 100], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ [50, 100]. Plot them on two separate panels. In 3–5 sentences, explain:

- What the MTM profile shape reveals about the portfolio's net directionality and curvature.
- Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
- What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the $50–$100 range.

```{r q1}

#Initializing dividend as 0 

#Since we are dealing with futures contracts we will not be using BSM, instead we will use Black 76
#The numerical difference is that our cost of carry (r - d) will be set to 0, so we will initialize b = 0 and remove b = r - d
b = 0 


#This function will be used to calculate delta using the numerical method 
numericaldelta <- function(price, type, K, T2M, sigma){
  step.delta = 1 
  
  p.down <- RTL::GBSOption(price - step.delta, K , T2M, r , b = 0, sigma = sigma, type = type)$price
  p.current <- RTL::GBSOption(price, K , T2M, r , b = 0, sigma = sigma, type = type)$price
  p.up <-RTL::GBSOption(price + step.delta, K , T2M, r , b = 0, sigma = sigma, type = type)$price

  delta.numerical <- (p.up - p.down) / 2 * step.delta
  delta.numerical
  }

  

numericalgamma <- function(price, type, K, T2M, sigma){
  step.gamma = 1 
  
  p.down <- RTL::GBSOption(price - step.gamma, K , T2M, r , b = 0, sigma = sigma, type = type)$price
  p.current <- RTL::GBSOption(price, K , T2M, r , b = 0, sigma = sigma, type = type)$price
  p.up <-RTL::GBSOption(price + step.gamma, K , T2M, r , b = 0, sigma = sigma, type = type)$price

  gamma.numerical <- (p.up - 2 * p.current + p.down) / step.gamma ^ 2
  gamma.numerical
  }



numericalvega <- function(price, type, K, T2M, sigma){
  step.vega = 0.01  
  
  p.down <- RTL::GBSOption(price, K , T2M, r , b = 0, sigma = sigma - step.vega, type = type)$price
  p.current <- RTL::GBSOption(price, K , T2M, r , b = 0, sigma = sigma, type = type)$price
  p.up <-RTL::GBSOption(price, K , T2M, r , b = 0, sigma = sigma + step.vega, type = type)$price

  vega.numerical <- (p.up - p.down) / 2 * step.vega * 100
  vega.numerical
  }



numericaltheta <- function(price, type, K, T2M, sigma){
  step.theta = 1/365
  
  p.down <- RTL::GBSOption(price, K , T2M - step.theta, r , b = 0, sigma = sigma, type = type)$price
  p.current <- RTL::GBSOption(price, K , T2M, r , b = 0, sigma = sigma, type = type)$price
  
  theta.numerical <- p.down - p.current
  theta.numerical
  }




#We will duplicate port to port3 and add the greeks to it, computing them numerically for F_0 = 73 
port3 = port %>% mutate(
  F_0 = F_0,
  delta = mapply(numericaldelta, F_0, type, K, T2M, sigma),
  gamma = mapply(numericalgamma, F_0, type, K, T2M, sigma), 
  vega = mapply(numericalvega, F_0, type, K, T2M, sigma), 
  theta = mapply(numericaltheta, F_0, type, K, T2M, sigma))



#For each price we need to show Delta, Gamma, Vega, Theta for each of the three positions

#All prices 50-100 
prices <- tibble(price = seq(50,100))

#Cross join so we have all combinations 
port2 <- cross_join(prices, port)







#Adding greeks
port2 <- port2 %>% mutate(delta = mapply(numericaldelta, price, type, K, T2M, sigma), 
                          gamma = mapply(numericalgamma, price, type, K, T2M, sigma), 
                          vega = mapply(numericalvega, price, type, K, T2M, sigma), 
                          theta = mapply(numericaltheta, price, type, K, T2M, sigma))




#Create plots showing each Greek vs Futures price over f = 50-100, with one line per position
#Three lines for each plot, each plot will represent one of the greeks, four plots total 
#We will express each greek as a total PL impact on the position to capture the directionality and size of the position, even though they all represent 2000 contracts 




port2 %>% 
  ggplot() +
  geom_line(aes(x = price, y = delta*(n*mult), color = pos))
  
port2 %>% 
  ggplot() +
  geom_line(aes(x = price, y = gamma*(n*mult), color = pos))

port2 %>% 
  ggplot() +
  geom_line(aes(x = price, y = vega*(n*mult), color = pos))

port2 %>% 
  ggplot() +
  geom_line(aes(x = price, y = theta*(n*mult), color = pos))

#Will need to double check if these plots look right 




#d. Portfolio mark to market value and gamma 


#Going to just make a gbswrapper function (like class notes) to simplify mapply implementation
#I'll make it just return price to start but can modify later if needed
gbswrapper <- function(price, K, T2M, r, b, sigma, type) {
  RTL::GBSOption(price, K, T2M, r, b, sigma = sigma, type = type)$price
}



#Calculating the values for all options positions 
port2 <- port2 %>% 
  mutate(
    optvalue = mapply(gbswrapper, price, K, T2M, r, b = 0, sigma, type), 
    posvalue = optvalue*n*mult)



#Creating portfolio value combining the individual positions
port4 <- port2 %>%
  group_by(price) %>%
  summarise(portvalue = sum(posvalue)) %>% 
  ungroup()

port4 %>% ggplot() +
  geom_line(aes(x = price, y = portvalue))





port5 <- port2 %>%
  mutate(gammavalue = gamma*n*mult) %>%
  group_by(price) %>%
  summarise(portgamma = sum(gammavalue))

port5 %>% ggplot() +
  geom_line(aes(x = price, y = portgamma))



```

# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2}



#Sum greeks across positions to show portfolio level greeks 
#Need to summarize the greeks for all price levels

#Numbers too high, need to check how I'm multiplying them by mult and n 

port6 <- port2 %>%
  mutate(deltavalue = delta*n*mult) %>%
  mutate(gammavalue = gamma*n*mult) %>%
  mutate(vegavalue = vega*n*mult) %>%
  mutate(thetavalue = theta*n*mult) %>%
  group_by(price) %>%
  summarise(portdelta = sum(deltavalue),
            portgamma = sum(gammavalue),
            portvega = sum(vegavalue),
            porttheta = sum(thetavalue))




#port6 %>% ggplot(aes(x=price)) +
#  geom_line(aes(y = portdelta, color = "Delta")) + 
#  geom_line(aes(y = portgamma, color = "Gamma")) + 
#  geom_line(aes(y = portvega, color = "Vega")) + 
#  geom_line(aes(y = porttheta, color = "Theta"))  



#I am interpreting this as just showing the portfolio greeks at F_0 which is 73 
#port6 has the portfolio greeks for all prices 
port6 %>% 
  filter(price == F_0) %>%
  gt() %>%
  tab_header(
    title = "Portfolio Greeks",
  )

#Have determined through changing the price range, and considering the nature of the portfolio that delta will never be positive, it is a delta negative portfolio 




```

# Question 3 (8 points)

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

+ WTI futures: F moves from **73.00** to **90.00** (+$17.00/bbl)
+ Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
+ Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1. **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2. **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3. **Unexplained residual**: Actual − Attributed.
4. **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3}














```

# Question 4 (4 points)

The Iran war scenario (ΔF = +$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ [–20, +20] in $1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4}







```

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

+ Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the $50–$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
+ Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
+ Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
+ Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
+ Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

*Your answer here.*
